---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Brandon Tran\
MATH 154

# Work in Progress for Final Project

## Generating Test Data
For testing purposes, we will utilize the dataset for San Francisco.  Note that we will have to omit any observations with $NA$ entries for any one of `lat`, `lng`, `date`, `time`, and `subject_race`.

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
```

```{r eval=FALSE}
sf <- readRDS(file = url('https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_san_francisco_2019_08_13.rds'))
sf.api <- sf %>%
  select(c(date, time, lat, lng, subject_race)) %>%
  drop_na()
```

## Functions to Classify Times as Day or Night
First, we'll write a few helper functions that utilize `POSIXct`, `hms`, and the data itself in order to first determine the sunrise and sunset times for a given day and coordinates, then return a boolean evaluating to `True` if the observed time was during the day and `False` if it occurred at night.

```{r, warning=FALSE, message=FALSE}
library(StreamMetabolism)
library(hms)
```

```{r}
get_time <- function(lat, long, date) {
  #Takes lat, long, time, and date as input and classifies TRUE if the time is at day and FALSE if at night.
  
  sunrise <- sunrise.set(lat, long, date)$sunrise
  attributes(sunrise)$tzone <- 'America/Los_Angeles'
  sunrise <- as_hms(sunrise)
  
  sunset <- sunrise.set(lat, long, date)$sunset
  attributes(sunset)$tzone <- 'America/Los_Angeles'
  sunset <- as_hms(sunset)
  
  return(c(sunrise, sunset))
}

classify <- function(rise.set, time) {
  #Takes in a vector with sunrise and sunset times, returning True if the time is during daylight.
  
  if (time < rise.set[2] && time >= rise.set[1])
    return(TRUE)
  
  else
    return(FALSE)
}
```

Next, we'll write a second function that will apply our helper function to any given dataframe, appending the booleans as a new column variable.

```{r}
mutateClass <- function(df) {
  #Takes a dataframe and adds the boolean day/night classification found via `classify` as a new column.
  
  classifications <- c()
  timetoset <- c()
  
  for (i in 1:nrow(df)) {
    time <- get_time(df$lat[i], df$lng[i], df$date[i])
    timetoset[i] <- (time[2] - df$time[i])/60
    classifications[i] <- classify(time, df$time[i])
  }
  
  df <- df %>%
    mutate(daytime = classifications) %>%
    mutate(minsto.set = timetoset)
}
```

This function can now be used to generate `True` and `False` classifications for our new variable, `daytime`.

## Putting it All Together
We now have some data, and we have functions that will classify an observation (really, a traffic stop) as during or not during the daytime.  We now want to parse data together to form an appropriate sample of the State of California, which will put us in a good place for both our Shiny visualization and the statistical testing for signficance.  Note that for some datasets (namely, Los Angeles, San Diego, Oakland, and San Jose), `lat` and `long` are not included.  Luckily, we only need the location in order to calculate the exact sunrise/sunset times, and this can be generalized to each city.  Our function will include functionality to create the necessary `lat` and `long` variables for datasets without such information.

```{r}
getCityCoords <- function(link) {
  #Takes a link to data and assigns appropriate lat and lng coordinates.
  
  if (grepl('los_angeles', link))
    return(c(34.052235, -118.243683))
  
  if (grepl('san_diego', link))
    return(c(32.715736, -117.161087))
  
  if (grepl('oakland', link))
    return(c(37.804363, -122.271111))
  
  if (grepl('san_jose', link))
    return(c(37.335480, -121.893028))
}

getData <- function(link) {
  #Takes a link to data and configures it as necessary.  Samples n/10000 traffic stops.  Returns a mutated `daytime` variable.
  
  set.seed(47)
  
  raw <- readRDS(file = url(link))
  
  if (! 'lat' %in% colnames(raw)) {
    raw <- raw %>%
      mutate(lat = getCityCoords(link)[1]) %>%
      mutate(lng = getCityCoords(link)[2])
  }

  clean <- raw %>%
    select(c(date, time, lat, lng, subject_race)) %>%
    drop_na() %>%
    sample_n(nrow(raw)/100)
  
  return(mutateClass(clean))
}
```

We now have a function, `getData`, that will effectively scrape all of the necessary datasets and generate the samples and daytime booleans that will be used to conduct our statistical analyses and visualizations.  The following function calls will generate individual datasets for each city.

```{r eval=FALSE}
sf <- getData('https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_san_francisco_2019_08_13.rds')
ok <- getData('https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_oakland_2019_08_13.rds')
sj <- getData('https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_san_jose_2019_08_13.rds')
bf <- getData('https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_bakersfield_2019_08_13.rds')
la <- getData('https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_los_angeles_2019_08_13.rds')
sd <- getData('https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_san_diego_2019_08_13.rds')
```

Our final goal is to create one large dataset containing all of the data samples.

```{r}
parseData <- function(links, city.codes) {
  #Takes in a list of URLs and city codes and merges the datasets that are scraped from the sources.
  
  master.frame <- getData(links[1])
  master.frame <- master.frame %>% mutate(city = city.codes[1])
  
  for (i in 2:length(links)) {
    data <- getData(links[i])
    data <- data %>% mutate(city = city.codes[i])
    master.frame <- full_join(master.frame, data)
  }
  
  return(master.frame)
}
```

And, we're done!  Running the following code will generate the dataset that we need:

```{r eval=FALSE}
links <- c("https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_san_francisco_2019_08_13.rds",
           "https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_oakland_2019_08_13.rds",
           "https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_san_jose_2019_08_13.rds",
           "https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_bakersfield_2019_08_13.rds",
           "https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_los_angeles_2019_08_13.rds",
           "https://stacks.stanford.edu/file/druid:hp256wp2687/hp256wp2687_ca_san_diego_2019_08_13.rds")
city.codes <- c('sf', 'ok', 'sj', 'bf', 'la', 'sd')

ca.df <- parseData(links, city.codes)
```

To save a local copy of this dataframe, run the following code chunk.  A copy has been saved to the repo in the `Data` folder.

```{r eval=FALSE}
save(ca.df, file='cadata.Rda')
```

## Some Data Analysis (Section Under Review)
So, what do we find?  Is there a statistically significant difference between the proportion of stops of members of a certain race during the day versus during the night?  To analyze this, we'll use a $\chi^2$ test.

```{r, eval=FALSE}
#In General
chisq.test(ca.df$subject_race, ca.df$daytime)

#Local Inference for API
ca.api <- ca.df %>%
  filter(subject_race == "asian/pacific islander" | subject_race == "white")

chisq.test(ca.api$subject_race, ca.api$daytime)

#Local Inference for Black
ca.black <- ca.df %>%
  filter(subject_race == "black" | subject_race == "white")

chisq.test(ca.black$subject_race, ca.black$daytime)

#Local Inference for Hispanic
ca.hisp <- ca.df %>%
  filter(subject_race == "hispanic" | subject_race == "white")

chisq.test(ca.hisp$subject_race, ca.hisp$daytime)
```

The answer is yes: With five degrees of freedom, we get a p-value of $p = 2.2 \times 10^{-16}$, suggesting that race and time of stop are not independent of each other.  However, via local inference under multiple testing considerations (including the utilization of Bonferroni's inequality), we find that contrary to our expectation, the disparity between daytime and nighttime arrests is really only significant when the violator is white.


## Official Blurb on Statistical Analysis
We performed a $\chi^{2}$ test for Independence in order to evaluate the null hypothesis: that `race` and `time of arrest` are independent of each other, where `time of arrest` refers to whether or not a stop is before or after sunset.  With $5$ degrees of freedom, the test yielded a $p$-value of $p = 2.2 \times 10^{-16}$, which is less than $\alpha = .05$, thus allowing us to reject the null and conlcude that the two variables are, in fact, dependent.

In more colloquial terms, we disproved the assumption that knowing what time of day it is has no effect on one's guess of the race of a random driver.  Rather, a relationship does, in fact, exist here: If given the time of day, we should guess a certain race moreso than the others when considering a random driver.

Which race, specifically, should we guess?  Unforunately, our test does not tell us this, but some local analysis will allow us to test some of our predispositions.  Specifically, we will test `white` against `Asian/Pacific Islander`, `Black`, and `Hispanic`.  For the sake of technicality: Under Bonferroni's Inequality, we need the sum of our $p$-values to be less than $\alpha = .05$ for significance.  When comparing each of the above races against `white`, we gain small $p$-values that sum to less than $\alpha = .05$, suggesting that the distribution of `white` traffic stops differs from the rest.

Visualizations will aid in the interpretation of these results.



